### 2025-09-22 paper/sec/introduction.tex
Initial progress on target idioms.

NLP aspect of the problem is potentially a big problem space in itself. We will restrict interest to certain
idiomatic uses of natural language in making/justifying scientific claims. Table~\ref{tab:fluid_examples}

>>>>

Table~\ref{tab:fluid_examples} summarises the natural language idioms we study in this paper. When using
state-of-the-art models like gpt4o and gpt5, our system is able to resolve basic table lookups of direct
numerical values, as well as computations of percentages, averages, minima and maxima, and totals, each mapped
to a corresponding aggregation over the source dataset. For example, phrases such as ``the Energy Sector
accounts for 52.80\% of total emissions'' or ``the expected projected methane emissions for 2030 is 13.5'' are
interpreted as division and averaging operations over the relevant data values. Similarly, ``recorded its
highest emissions in 2030'' is interpreted as a \kw{maximumBy} query, while rank-based statements such as
``CNN gives the lowest accuracy'' are interpreted as ranking computations.

We also consider \emph{trend} expressions, which include relative or comparative natural language phrases
describing how some property evolves over time. These include single comparisons (e.g.~``training time growing
from 67 to 106 seconds'') and universally quantified comparisons (e.g.~``SVM outperformed all other
kernels''). Such idioms are mapped to higher-level operators like \kw{trendWord} which are parameterised on
helper functions such as \kw{growShrink} and \kw{betterWorse} which maps comparisons to appropriate natural
language phrases. Taken together, these categories cover a substantial portion of the numerical reasoning
idioms found in empirical scientific writing.

Notably, the current coverage omits several common linguistic forms. We have yet to study the ability of LLMs
to interpret approximate quantitative terms like \textbf{around 50\%} or \textbf{roughly 100 instances}, nor
interval-based descriptions such as \textbf{between 30 and 40}\% or \textbf{within 5–10 seconds}. We have no
reason for thinking these will present specific difficulties, however. So-called \emph{graded} modal
adjectives which express comparatives and other adjectives in nuanced or qualified ways -- like
\textbf{slightly better} and \textbf{significantly higher} -- are likely to be more
challenging~\cite{lassiter17}, especially when combined with trends over time, as in \textbf{steady increase}
or \textbf{sharp decline}. Distributional descriptors like ``most models'', ``the majority of cases'', or
``usually performs better than'' present significant challenges to formal interpretation because of their
reliance on non-standard quantifiers. Extending coverage to these idioms would substantially deepen our tool's
ability to bridge natural language reporting with interpretation in terms of the underlying dataset. We
discuss this further in \secref{future-work}.
